Our Galactic Midplane Is Probably Wrong, and Not by a Little

### 2019-02-11 paper/ms.tex
Rewrite abstract

In a completely axisymmetric potential, the notion of a global Galactic
midplane is well defined. However, since the Milky Way is not completely
axisymmetric, the global Galactic midplane is to some level of precision not
well defined. The local Galactic midplane has been found by determining the
height at which the number counts of stars reaches a maximum. This local
midplane is then used as the global midplane, an assumption which would be
justified in a completely axisymmetric potential. Assuming a global Galactic
midplane exists, we show in three cosmological zoom-in simulations from the
FIRE collaboration that local determinations of the midplane can vary by up to
a thin disk scale height. The effect is more pronounced in galaxies with less
quiescent histories (e.g. Sagittarius-like encounters). We explore one
consequence of such an inaccuracy of the midplane in the context of orbital
actions. Using standard orbit integration techniques, we show that midplane
inaccuracy induces artifical time evolution in the vertical action that makes
it virtually unusable for individual stars in the disk. The situation is less
dire for angular momentum and radial action determinations, though we also
show that inaccuracies in the solar radius can have a similar effect on the
latter. As a result, we question the use of actions as dynamical invariants in
the Milky Way disk.

>>>>

We propose a new type of error in the computation of stellar actions. This
error is induced by the fact that the local Galactic midplane varies across
the disk. As a result, when we make 6D phase space measurements of stars that
are non-local (i.e. far away), we are observing those stars with a systematic
offset in $z$. We show that this systematic offset induces time evolution in
the observed actions which we interpret as an error. We note that this error
is non-Gaussian and bimodal, with neither of the modes peaking on the true
action value. A midplane error of $\sim30\,\pc$ and $\sim225\,\pc$ is
necessary to induce a $100\,\%$ error in the vertical action $J_z$ for thin
and thick disk orbits, respectively. Using cosmological zoom-in simulations
from the FIRE collaboration, we show that the midplane can vary by
$\sim100\,\pc$ around the Solar circle, indicating that such an effect is
important for understanding the thin and thick disk. Our discussion is also
directly applicable to variations in the local standard of rest. The variation
of the midplane must be taken into account when performing dynamic modelling
across large regions of the disk accessible with the {\em Gaia} mission.
